---
title: "Using Color in R"
author: "Earl F. Glynn, Principal Programmer/Analyst, UMKC Center for Health Insights"
highlighter: highlight.js
hitheme: tomorrow
job: Kansas City R Users Group, 4 January 2014
mode: selfcontained
subtitle: null
framework: io2012
widgets: mathjax
---

```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 100)
knitr::opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knit_hooks$set(plot = knitr:::hook_plot_html)
```

<!-- Set background color for slide 1 in libraries/frameworks/io2012/slidify.css -->

## Key ideas
* Color Spaces
  * RGB Additive Color Space
  * RGB Color Names
* R Color Basics
  * Color Definition
  * Palette Types and RColorBrewer
  * Points, Axes, Text, Lines, Legends, Background
* Color Tips
  * Image Processing
  * Color Deficiency/Blindness
  * Why don't screen colors match printout?
  * Presentations

---

## Color Spaces
* [RGB](http://en.wikipedia.org/wiki/RGB_color_space) Red-Green-Blue

* [HSV](http://en.wikipedia.org/wiki/HSL_and_HSV) Hue-Saturation-Value

* [HLS](http://en.wikipedia.org/wiki/HSL_and_HSV) Hue-Lightness-Saturation

* [LAB](http://en.wikipedia.org/wiki/Lab_color_space) coordinates in CIE LAB space

* R Package: [help(package=colorspace)](http://cran.r-project.org/web/packages/colorspace/colorspace.pdf)

[Why did you learn the wrong primary colors in elementary school?](http://www.greatreality.com/ColorPrimary.htm)

---

## RGB Additive Color Space
<img class=center src=assets/img/RGB-Color-Cube.png height='80%' />

---

## RGB Additive Color Space
0.0 to 1.0, decimal 0 to 255, hexadecimal 00 to FF

```{r}
rgb(1,0,0)   # red
rgb(0,1,0)   # green
rgb(0,0,1)   # blue
```

---

## RGB Additive Color Space
```{r}
rgb(1,0,1)   # magenta = red   + blue
rgb(1,1,0)   # yellow  = red   + green
rgb(0,1,1)   # cyan    = green + blue
```

---

## RGB Additive Color Space
```{r}
rgb(1,0,0)    # red
rgb(255,0,0, maxColorValue=255)
col2rgb(c("black", "red", "green", "blue", "yellow", "white"))
```

---

## RGB Color Names:  `colors()`
Color names and RGB values also in file
C:\Program Files\R\R-3.0.2\etc\rgb.txt

```{r}
# colors() or colours()
length(colors())
head(colors())
tail(colours())
```

---

## RGB Color Names
```{r}
colors()[grep("red", colors())]   # colors with "red"
head(colors()[grep("blue", colors())], 25)  # 25 of 66 "blue" colors
```
---

## RGB Color Names
7-page Color Chart of colors()

[https://github.com/EarlGlynn/colorchart/raw/master/ColorChart.pdf](https://github.com/EarlGlynn/colorchart/raw/master/ColorChart.pdf)

<img class=center src=assets/img/R-Color-Names-RGB.png height='30%' />

Details showing how to create chart in R:  [Color Chart in R](https://github.com/EarlGlynn/colorchart/wiki/Color-Chart-in-R)

---

## RGB Color Names

__3D colors() Example__

From Example 6a in `?scatterplot3d`  help in `library(scatterplot3d)`
```{r ,echo=FALSE,fig,fig.height=4.5,fig.width=5,fig.cap="colors() scatterplot3d"}
library(scatterplot3d)
## example 6; by Martin Maechler
cubedraw <- function(res3d, min = 0, max = 255, cex = 2, text. = FALSE)
{
  ## Purpose: Draw nice cube with corners
  cube01 <- rbind(c(0,0,1), 0, c(1,0,0), c(1,1,0), 1, c(0,1,1), # < 6 outer
                  c(1,0,1), c(0,1,0)) # <- "inner": fore- & back-ground
  cub <- min + (max-min)* cube01
  ## visibile corners + lines:
  res3d$points3d(cub[c(1:6,1,7,3,7,5) ,], cex = cex, type = 'b', lty = 1)
  ## hidden corner + lines
  res3d$points3d(cub[c(2,8,4,8,6),     ], cex = cex, type = 'b', lty = 3)
  if(text.)## debug
      text(res3d$xyz.convert(cub), labels=1:nrow(cub), col='tomato', cex=2)
}
## 6 a) The named colors in R, i.e. colors()
cc <- colors()
crgb <- t(col2rgb(cc))
par(xpd = TRUE)
rr <- scatterplot3d(crgb, color = cc, box = FALSE, angle = 24,
    mar = c(3,4,0,0),   # Added to example
    xlim = c(-50, 300), ylim = c(-50, 300), zlim = c(-50, 300))
cubedraw(rr)
```

---
## R Color Basics:  Color Definition

A color in R can be specified in three ways:

* color name

* hex constant #000000 through #FFFFFF  (a 24-bit color)

* index into palette, which is vector of color names and/or hex constants

---
## R Color Basics:  Default Palette

```{r,fig.height=2,fig.width=5}
palette()   # show default palette
par(mar=c(1,1,1,1))  # Reduce margins
barplot(rep(1,8), yaxt="n", col=1:8)
```

To use default palette if not currently the defined palette:

`barplot(rep(1,8), yaxt="n", col=palette("default"))`

---
## R Palette:  "Rainbow"

```{r,fig.height=2,fig.width=5}
N <- 20
palette(rainbow(N))   # Define N-color rainbow palette
palette()             # Show colors in current palette
par(mar=c(1,1,1,1))   # Reduce margins
barplot(rep(1,N), yaxt="n", col=1:N)
```

---
## R Palette:  Shades of Grey Palette

Function spelled either `grey()` or `gray()`

```{r,fig.height=2,fig.width=5}
N <- 16                           # N = Number of grey levels
0:(N-1) / (N-1)                   # vector with values from 0 to 1
palette(grey(0:(N-1) / (N-1)))    # Define N-shades of grey (256 is max)
palette()                         # Show colors in current palette
```

---
## R Palette:  Shades of Grey Palette

```{r,fig.height=2,fig.width=5}
par(mar=c(1,1,1,1))                     # Reduce margins
barplot(rep(1,3*N), yaxt="n", col=1:N)  # Three cycles
```

Plotting 3*N bars with palette of N shades of grey results in color recycling.
R recycles colors whenever there are more objects to plot than colors.

---

## R Palette:  Other N-Color Palettes

__heat.colors()__

```{r,fig.height=1,fig.width=5}
par(mar=rep(0,4))    # No margins
barplot(rep(1,16), yaxt="n", col=heat.colors(16))
```

__cm.colors()__    *cyan-magenta*

```{r,fig.height=1,fig.width=5}
par(mar=rep(0,4))    # No margins
barplot(rep(1,16), yaxt="n", col=cm.colors(16))
```

---
## R Palette:  Other N-Color Palettes

__topo.colors()__

```{r,fig.height=1,fig.width=5}
par(mar=rep(0,4))    # No margins
barplot(rep(1,16), yaxt="n", col=topo.colors(16))
```

__terrain.colors()__

```{r,fig.height=1,fig.width=5}
par(mar=rep(0,4))    # No margins
barplot(rep(1,16), yaxt="n", col=terrain.colors(16))
```

---

## R Palette Function:  colorRampPalette()

Use `colorRampPalette()` to define function that creates palette.

```{r,fig.height=4.5,fig.width=6}
z <- outer(1:20, 1:20, function(x,y) sin(sqrt(x*y)/3))
PaletteFunction <- colorRampPalette(c("red", "white", "blue"), space = "Lab")
filled.contour(z, col = PaletteFunction(20))
```

---
## Palette Types

__Qualitative Palette__
* All colors have same perceptual weight/importance
* Typical application:  bar plot

__Sequential Palette__
* For coding numerical information in a range
* Typical application:  heat map

__Diverging Palette__
* Like sequential palette but middle neutral value between two extremes

Details:
* [Choosing Color Palettes for Statistical Graphics](http://epub.wu.ac.at/1404/1/document.pdf)
* [Choosing Colour palettes. Part I: Introduction](http://blog.ggplot2.org/post/23995319650/choosing-colour-palettes-part-i-introduction) (ggplot2 examples)

---
## R Palettes:  RColorBrewer Package

```{r}
library(RColorBrewer)
table(brewer.pal.info$category)
brewer.pal.info[brewer.pal.info$category == "qual",]
```

---
## RColorBrewer: Qualitative Palettes

```{r,fig.height=4.5,fig.width=6}
par(mar=c(0,4,0,0))  # Smaller margins than default
display.brewer.all(type="qual")
```

---
## RColorBrewer: Sequential Palettes

```{r,fig.height=4.5,fig.width=6}
par(mar=c(0,4,0,0))  # Smaller margins than default
display.brewer.all(type="seq")
```

---
## RColorBrewer: Diverging Palettes

```{r,fig.height=4.5,fig.width=6}
par(mar=c(0,4,0,0))  # Smaller margins than default
display.brewer.all(type="div")
```

---

## R Color Basics:  Object color

Many R objects can take on different colors:

* Points
* Lines
* Axes
* Text
* Legends
* Background

```{r ,echo=FALSE}
palette("default")
```

---

## R Color Basics:  Points

```{r, fig.height=3,fig.width=3}
palette()[1:5]   # First five elements of default palette
x <- -2:2
y <- x^2
plot(x,y, col=1:5, pch=CIRCLE<-16, cex=2)
```
---

## R Color Basics:  Points

__Equivalent plot statements:__

assuming default palette

`plot(x,y, col=1:5, pch=CIRCLE<-16, cex=2)`

`plot(x,y, col=c("black", "red", "green3", "blue", "cyan"), pch=CIRCLE, cex=2)`

`plot(x,y, col=c("#000000", "#FF0000", "green3", 4, 5), pch=CIRCLE, cex=2)`

---

## R Color Basics:  Points, Axes, Text

```{r, echo=FALSE, fig.height=5,fig.width=5}
plot(0:10, 0:10, col=0:10, pch=CIRCLE<-16, axes=FALSE,
     main="Chart Title", col.main="blue", cex.main=2,
     xlab="X axis", ylab="Y axis", col.lab="blue", cex.lab=1.5)
mtext("Bottom",  BOTTOM<-1, col="red")   # Margin T
mtext("Left",    LEFT<-2,   col="green")
mtext("Top",     TOP<-3)
# Axes
axis(BOTTOM, col="red",   col.axis="black", cex.axis=1.5)
axis(LEFT,   col="green", col.axis="green", at=2*0:5, labels=paste(20*0:5),
     las=AXIS_LABEL_HORIZONTAL<-1)
```

See code on next slide.

---
## R Color Basics:  Points, Axes, Text

```{r, eval=FALSE}
# Points, Title, Axis Labels
plot(0:10, 0:10, col=0:10, pch=CIRCLE<-16, axes=FALSE,
     main="Chart Title", col.main="blue", cex.main=2,
     xlab="X axis", ylab="Y axis", col.lab="blue", cex.lab=1.5)

# Margin Text
mtext("Bottom",  BOTTOM<-1, col="red")
mtext("Left",    LEFT<-2,   col="green")

# Axes
axis(BOTTOM, col="red",   col.axis="black", cex.axis=1.5)
axis(LEFT,   col="green", col.axis="green", at=2*0:5, labels=paste(20*0:5),
     las=AXIS_LABEL_HORIZONTAL<-1)
```

---
## R Color Basics:  Lines and Legends

```{r, echo=FALSE, fig.height=5,fig.width=7}
library(RColorBrewer)

Time <- 0:120  # minutes

Types  <- c("solid", "dotted", "dashed")
Colors <- brewer.pal(3, "Dark2")

Periods <- data.frame(Period1=cos(2*pi*Time/120),
                      Period2=cos(2*pi*Time/ 90),
                      Period3=cos(2*pi*Time/150))

matplot(Periods, type="l", lwd=3,
        main="Expression Vs. Time", cex.main=1.5,
        xlab="Time[min]", ylab="Expression",
        col=Colors, lty=Types)
legend("top", c("120 min period", "90 min period", "150 min period"),
       col=Colors, lty=Types, lwd=3)
```

See code on next slide.

---
## R Color Basics:  Lines and Legends

```{r, eval=FALSE}
library(RColorBrewer)

Time <- 0:120  # minutes

Types  <- c("solid", "dotted", "dashed")
Colors <- brewer.pal(3, "Dark2")   # qualitative palette

Periods <- data.frame(Period1=cos(2*pi*Time/120),
                      Period2=cos(2*pi*Time/ 90),
                      Period3=cos(2*pi*Time/150))

matplot(Periods, type="l", lwd=3,
        main="Expression Vs. Time", cex.main=1.5,
        xlab="Time[min]", ylab="Expression",
        col=Colors, lty=Types)
legend("top", c("120 min period", "90 min period", "150 min period"),
       col=Colors, lty=Types, lwd=3)
```

---
## R Color Basics:  Background Color

```{r, fig.height=3,fig.width=3}
par(bg="wheat", mar=c(4,4,0,0))  # background color
x <- -2:2
y <- x^2
plot(x,y, col=1:5, pch=CIRCLE<-16, cex=2)
```

* Useful when copying and pasting metafile to PowerPoint since default is "transparent"

---
## Color Tip:  Image Processing

* Bioconductor package EBImage is an image processing toolbox for R.
* Supports `colormode` of `Grayscale` or `Color`.
* `channel` function handles color space conversions.
* Color colormode images have red, green, blue and possiby alpha channels.

<img class=center src=assets/img/EBImage-Segmentation.png height='100%' />

Segmentation of cells in *Introduction to EBImage* documentation.


---
## Color Tip:  Color Deficiency/Blindness

* About 1 in 12 have some sort of color deficiency.
* ~8% of men and ~0.5% of women are affected [according to Wikipedia](http://en.wikipedia.org/wiki/Color_blindness).

<img class=center src=assets/img/Ishihara-Tests.png height='70%' />

[Ishihara Color Blindess Tests from ishiharatest.blogspot.com](http://ishiharatest.blogspot.com/)

* What numbers do you see?  12 at the left.  29 or 70 at the right?
* [How the Color Deficient Person Sees the World](http://colorvisiontesting.com/what%20colorblind%20people%20see.htm)


---
## Color Deficiency/Blindness:  R dichromat package

```{r, fig.height=3,fig.width=6}
library(dichromat)
par(mfcol=c(1,2), mar=rep(0,4))
N <- 20
pie(rep(1,N), col=heat.colors(N))
pie(rep(1,N), col=dichromat(heat.colors(N)))
```

`dichromat` function collapses red-green color distinctions to approximate the effect of the two common forms of red-green color blindness: protanopia and deuteranopia

---
## Color Deficiency/Blindness:  R dichromat package

```{r, echo=FALSE, fig.height=5,fig.width=6}
library(dichromat)
par(mar=c(1,2,2,1))
layout(matrix(1:6,ncol=1))
image(1:10,1,matrix(1:10, ncol=1),
     col=colorschemes$BrowntoBlue.10,
     main="Brown to Blue (10)", axes=FALSE)
image(1:100,1,matrix(1:100 ,ncol=1),
     col=colorRampPalette(colorschemes$BrowntoBlue.10,space="Lab")(100),
     main="Brown to Blue Ramp", axes=FALSE)
image(1:10,1,matrix(1:10, ncol=1),
     col=dichromat(colorschemes$BrowntoBlue.10),
     main="Brown to Blue (10) -- deuteranopia", axes=FALSE)
image(1:12,1,matrix(1:12, ncol=1),col=colorschemes$Categorical.12,
     main="Categorical (12)", axes=FALSE)
image(1:12,1,matrix(1:12, ncol=1),
     col=dichromat(colorschemes$Categorical.12, "deutan"),
     main="Categorical (12) -- deuteranopia", axes=FALSE)
image(1:12,1,matrix(1:12,ncol=1),
     col=dichromat(colorschemes$Categorical.12, "protan"),
     main="Categorical (12) -- protanopia", axes=FALSE)
```

See code on next slide.

---
## Color Deficiency/Blindness:  R dichromat package

```{r, eval=FALSE}
library(dichromat)
par(mar=c(1,2,2,1))
layout(matrix(1:6,ncol=1))
image(1:10,1,matrix(1:10, ncol=1),
     col=colorschemes$BrowntoBlue.10,
     main="Brown to Blue (10)", axes=FALSE)
image(1:100,1,matrix(1:100 ,ncol=1),
     col=colorRampPalette(colorschemes$BrowntoBlue.10,space="Lab")(100),
     main="Brown to Blue Ramp", axes=FALSE)
...
image(1:12,1,matrix(1:12, ncol=1),col=colorschemes$Categorical.12,
     main="Categorical (12)", axes=FALSE)
image(1:12,1,matrix(1:12, ncol=1),
     col=dichromat(colorschemes$Categorical.12, "deutan"),
     main="Categorical (12) -- deuteranopia", axes=FALSE)
image(1:12,1,matrix(1:12,ncol=1),
     col=dichromat(colorschemes$Categorical.12, "protan"),
     main="Categorical (12) -- protanopia", axes=FALSE)
```

---
## Color Tip:  Why don't screen colors match printout?

* Every device has a gamut of colors that it can display.
* Color gamuts between devices often do not match.
  * How should a color be represented if it cannot be displayed on a device?
* Color conversions may not be perfect, e.g., RGB additive colors must be converted to CMYK (Cyan-Magenta-Yellow-Black) subtractive colors for printing.
* Paper/ink differences

---
## Color Tip:  Presentations/Graphics

* Avoid unnecessary use of color.
* Use bright colors with small graphics to make them stand out.
* Be consistent in use of color.
* Do NOT use color as only attribute to show difference. E.g., consider color and line type.

Also see:  [Public Speaking:  Cool Color Commentary](http://www.public-speaking.org/public-speaking-color-article.htm)

---

## Color Resources

* [Wheels, Pyramids, and Spinning Tops: The Scientific Approach to Color](http://www.lindahall.org/events_exhib/), Exhibit at Linda Hall Library through March 14

* [Choosing Color Palettes for Statistical Graphics](http://epub.wu.ac.at/1404/1/document.pdf), Wirtschaftsuniversität Wien
* [Color Chart in R](https://github.com/EarlGlynn/colorchart/wiki/Color-Chart-in-R)
* [Colour for Presentation Graphics](http://www.r-project.org/conferences/DSC-2003/Proceedings/Ihaka.pdf), Proceedings of 3rd International Workshop on Distributed Statistical Computing
* [Using Color in Information Display Graphics](http://colorusage.arc.nasa.gov/), NASA
* [Using RColorBrewer to Colour Your Figures in R](http://www.r-bloggers.com/r-using-rcolorbrewer-to-colour-your-figures-in-r/), R-bloggers
* [Why Should Engineers and Scientists be Worried About Color?](http://www.research.ibm.com/people/l/lloydt/color/color.HTM), IBM




